Efficient FPGA Implementation of\\ Digit Parallel Online Arithmetic Operators

### 2014-05-28 FPT2014/bare_conf.tex
George feedback 2

\begin{abstract}
%\boldmath
Online arithmetic has been widely studied for ASIC implementations. It is designed to perform computations from the most significant digits, resulting in the ability to chain arithmetic operators together for low latency. Recent research also shows that digit parallel online operators can fail more gracefully when operating beyond the deterministic clocking region, in comparison to operators with conventional arithmetic. Unfortunately, the utilization of online arithmetic operators in the past has required a large area overhead for FPGA implementation. In this paper, we propose novel approaches to implement key primitives of online arithmetic, such as adders and multipliers, efficiently on modern FPGAs with 6-input LUTs. The FPGA carry resources are used for both area reduction and performance improvements. We demonstrate experimentally that in comparison to a direct implementation, the proposed architectures achieve over $67\%$ Slice savings and over $1.2\times$ speed-ups for adders, and more than $3\times$ area savings and $1.5\times$ speed-ups for multipliers.

@@

\section{Introduction}\label{Sec:Intro}
In the conventional form of computer arithmetic, the computation results are generated either from the least significant digit (LSD), e.g. addition and multiplication, or from the most significant digit (MSD), e.g. division and square root. This inconsistency in computing directions will potentially result in large latency when propagating data among different operations. Online arithmetic was designed to solve this problem \cite{Ercegovac_OnlineOverview,Ercegovac_Book}. With online arithmetic, both inputs and outputs are processed in a MSD-first manner. This enables parallelism among multiple operations, and the overall computation latency can be significantly reduced. A brief overview of online arithmetic is given in Section~\ref{Sec:Background}.

Recent research has demonstrated another key advantage of online arithmetic: that it allows graceful degradation when operating beyond the deterministic clocking region, i.e. under overclocking \cite{SKDAC14_REVIEW}. This stream of research stems from the inspiration that for certain applications, releasing the absolute accuracy requirements could lead to significant performance improvements. However, traditional forms of computer arithmetic do not fail gracefully when timing violation happens, because timing errors tend to affect MSDs of the results. In comparison, online arithmetic is more ``overclocking friendly'', as timing errors only occur at LSDs of the results.

Despite its attractiveness, the usage of online arithmetic on existing FPGAs is still limited due to a large area overhead. For standard arithmetic, it is difficult to efficiently map arithmetic circuits solely onto look-up-tables (LUTs), the fundamental building blocks of FPGAs. As a result, both main commercial FPGA vendors introduced dedicated carry logic into the logic block to reduce the carry propagation delay by over one order of magnitude~\cite{Virtex6}. Significant work explores methods that can effectively map circuits to the carry logic for performance improvements~\cite{FPL10_FPGA_CarryChain,Prefix_CarryChain}, or which proposes alternative carry logic for FPGAs~\cite{FPGA_CarryChain_New1,FPGA_CarryChain_New2}. Furthermore, hard DSP blocks are also included in modern architectures to improve the area and performance of multiplications and multiply-accumulates~\cite{XilinxDSP}. However, most of these approaches are designed to accelerate computations with the conventional form of arithmetic.

For online arithmetic, two research directions have been investigated. First, although there has been previous research about FPGA implementations of online arithmetic, the work only focused on online adders; a brief review of relevant works in this area is presented in Section~\ref{Sec:OA_FPGA}. Second, there are existing approaches to build general purpose online adders are specifically for low cost FPGAs with 4-input LUTs and 2 LUTs within a Slice, such as the Altera Cyclone III \cite{AlteraCyclone} and Xilinx Spartan~3 series \cite{XilinxSpartan}. However, as will be discussed in Section~\ref{Sec:OA_FPGA}, these approaches cannot be directly applied on many modern FPGA with 6-input LUTs and 4 LUTs in a Slice, such as the Xilinx Virtex series FPGAs and all Xilinx 7 series FPGAs~\cite{Virtex7}. Moreover, the move to the 6-LUT on the basic architecture opens new opportunities for efficient implementation.
%However, the implementations of other key arithmetic primitives, such as online multipliers, lack exploration.

In this paper, both issues will be addressed. In Section~\ref{Sec:OA_FPGA}, we propose a novel approach of mapping the digit parallel online adder onto FPGA with 6-LUTs. The mapping is based on the fast carry logic, and the available resources within a Slice can be fully utilized. In Section~\ref{Sec:OM_FPGA}, we optimize the online multiplication algorithm to yield an efficient FPGA implementation of digit parallel online multiplier. In addition, the resource usage of both operators is given. We demonstrate experimentally on Xilinx Virtex-6 FPGAs that the proposed designs achieve significant area reduction and performance gain over their original implementations. For online adders, our method achieves over $67\%$ Slice savings and $1.2\times$ frequency speed-ups in comparison to a direct implementation. For online multipliers, the LUTs and Slice utilization of the proposed new architecture drops over $2.4\times$ and $3.2\times$, with a frequency speed-up more than $1.5\times$ in comparison to a direct implementation. In summary, the main contributions of this paper are an efficient mapping of online adders and multipliers to modern FPGAs. (to be completed)
% by fully utilizing the fast carry logic for .

@@

Online arithmetic has been widely used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, of which the data flow is illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required and $\delta$ is called ``online delay''. Notice that $\delta$ is normally a small constant, which is independent of the precision in a given operation. For ease of discussion, for the rest of this paper, the input data is normalized to fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.
%
\begin{figure}[tbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.pdf}
  \vspace{-4ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-1ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible with the employment of the redundant number system. Normally there are 2 most commonly used redundant number representation: carry-save (CS) \cite{CSadder} and signed-digit (SD) \cite{RedundantNumber}. For instance with SD representation, each digit is represented using a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. In comparison, the standard non-redundant representation only uses a digit set $\{0,\cdots,r-1\}$. Thus a standard number corresponds to several possible redundant representations. For example, the two's complement number $0.011$ can be represented in the binary online form as $0.1\overline{1}1$, $0.10\overline{1}$ and $0.011$ among many other possible representations.

Due to the redundancy, the MSDs of the result can be calculated using partial information from both inputs. Then the value of the number can be revised using the subsequent digits, because each number embodies multiple representations.

\subsection{Binary Online Addition}
Adders serve as a critical building block for arithmetic operations. To perform digit-parallel online addition, a redundant adder can be used directly. The structure diagram of an online adder with all signals represented with SD numbers is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module 3:2 denotes a 3:2 compressor, which takes 3 inputs and generates 2 outputs. Normally it can be implemented on the basis of a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labelled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only 2 FA delays for any operand word-length, with the cost of one extra FA for each digit of operands. This makes the online adder suitable for building up more complex arithmetic operators such as multipliers to accelerate the sum of partial products \cite{RedundantMult_1987}.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder2.pdf}
    \vspace{-6ex}
	\caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. Module 3:2 denotes the 3:2 compressor.}
    %\vspace{2ex}
	\label{Fig:Radix2SD_adder}
\end{figure}

@@

\subsection{Binary Online Multiplication}
Multiplication is another key arithmetic operator. Typically, online multiplication is performed in a recursive digit-serial manner, as illustrated in Algorithm~\ref{Algorithm:OnlineMult} \cite{Ercegovac_OnlineMult} where both inputs and outputs are $N$-digit numbers as given in~(\ref{Eq:Online_Operands}). For a given iteration $j$, the product digit $z_j$ is generated MSD-first through a selection function $sel()$. For any radix $r$ and chosen digit set, there exits an appropriate selection method and a value of $\delta$ which ensure convergence \cite{Ercegovac_OnlineMult}. As radix-2 is used most commonly in computer arithmetic, we keep $r=2$ throughout this paper with the corresponding redundant digit set $\{\overline{1},0,1\}$. In this case $sel()$ is given by (\ref{Eq:SelFunc_OM}) \cite{Oregon_OnlineNetwork}. Notice that the selection is based on 2 integer digits and 1 fractional digit of $W_{[j]}$.

@@

It is worth noting that the precision of the product $Z$ can be determined by directly truncating from the LSDs, since there is no carry propagations from the LSD to the MSD. This is useful especially when the multiplication product is used by successive computations, where only the first $N$ digits of the results are normally used. In comparison, the product is generated starting from the LSD with conventional arithmetic. Hence the successive operations cannot commence until the result with full-precision is generated. This feature of online arithmetic has inspired research about dynamically control of computation precision in numerous applications~\cite{Online_Truncation}.

@@

In Section~\ref{Sec:OM_FPGA}, we will first describe an optimized online multiplication algorithm and its FPGA implementation, which are designed specifically targeting on digit parallel operations. Then we provide a modified structure which only generates the most significant half of the product in order to achieve extra area savings.

@@

\section{Digit Parallel Online Adder on FPGAs}\label{Sec:OA_FPGA}
\subsection{Related Works}
There has been previous works about FPGA implementation of digit parallel online adders. From the literature, the existing approaches can be classified into three types:

\begin{enumerate}
    \item efficient mapping of the digit-parallel online adder onto sophisticated FPGAs resources \cite{FPT09RA,ASAP09RA};
    \item multiple operands addition by designing compressor trees based on bit counters \cite{TC13RA,CSMult_CounterArray};
    \item modifying existing FPGA architecture for more efficient digit-parallel online addition \cite{DAC07_CounterTree} and for specific applications \cite{FPL09_onlineCCM}.
\end{enumerate}

Type 2 examines the use of counters to sum $k$ $N$-digit numbers where $k>2$. Instead, we focus on two-operand online addition that performs $Z=X+Y$, as shown in Fig.~\ref{Fig:Radix2SD_adder}. In comparison to type 3, this paper attempts to obtain the best performance from mainstream FPGAs instead of suggesting ways to modifying the FPGA fabric.

Specifically in type 1, both works took advantage of the built-in carry resources in FPGAs. Conventionally the ASIC implementation of online adders is based on the 4:2 compressors, as outlined in Fig.~\ref{Fig:PreviousWork} within the gray background. However, directly applying this approach in the FPGAs could be less efficient. This is because there is no carry propagation between the 2 FAs within a 4:2 compressor, and the net delay between them can be large. Instead, Kamp et al \cite{FPT09RA} and Hormigo et al \cite{ASAP09RA} described very similar mapping techniques for online adders with SD representations and CS representations, respectively. They share the common idea that maps the logic block (\texttt{LB}) within the dotted circle in Fig.~\ref{Fig:PreviousWork}. In this case, the fast-carry logic in the FPGA can be employed, and the delay between the 2 FAs can be largely reduced.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder_42comp_Spartan2.pdf}
	\caption{Map the online adder onto Spartan FPGAs using the fast-carry resources. The grey background highlights the 4:2 compressor. Dotted circle indicates the logic block (LB) which can be mapped to the FPGA using the carry resources.}
	\label{Fig:PreviousWork}
\end{figure}

However, we notice the major limitations of both approaches that they only target on FPGAs with 4-input LUTs (4-LUT) and 2 LUTs within a logic Slice, such as the Xilinx Spartan series and the Altera Cyclone series. This is naturally reasonable because one \texttt{LB} can be mapped to a single Slice. Nevertheless, for FPGAs with 6-input LUTs (6-LUT) and 4 LUTs in a Slice, such as the Xilinx Virtex series and all Xilinx 7 series FPGAs, directly applying these approaches will result in either resource waste or logic fault. For instance, if two \texttt{LB}s are mapped to a Slice with 4 LUTs as seen in Fig.~\ref{Fig:PreWorkVirtex6}, the outputs of \texttt{LB1} will be faulty because the its carry input cannot be explicitly initialized. Due to the same reason, the output of \texttt{XORCY2} should not be used.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.45\textwidth]{./Figures/SDadder_FastCarry_PreviousWork.pdf}
	\caption{An example illustrating that direct applying approaches in \cite{FPT09RA} on a Virtex-6 FPGA will results in faulty outputs, because the carry input of \texttt{LB1} cannot be explicitly initialized.}
	\label{Fig:PreWorkVirtex6}
    %\vspace{-2ex}
\end{figure}

\subsection{Proposed Mapping Method}
To tackle this problem, we first modify the structure of the online adder to enable an efficient FPGA mapping. The structure of a 4-digit online adder is given as an example in Fig.~\ref{Fig:ModifiedOA}. In this equivalent structure, the first FA in each 4:2 compressor is split into 2 parts, which only generate carry and sum respectively. In this case they can be mapped individually on 2 LUTs. The second FA, which generates the outputs, is unchanged and can be implemented using the fast-carry logic.

@@

\begin{figure*}[tbp]
	\centering
	\includegraphics[width=.85\textwidth]{./Figures/SDadder_FPGA_New_4digit2.pdf}
	\caption{Modified structure of online adder. Left: an example of 4-digit online adder. The shaded part refers to the 2 logic blocks (LBs) that can be mapped onto 1 Slice. Right: one LB. The dotted box outlines the logic that can be mapped onto 1 LUT and the corresponding fast-carry logic.}
	\vspace{-1ex}	
	\label{Fig:ModifiedOA}
\end{figure*}

The detailed Slice mapping of the 2 \texttt{LB}s and the logic in each LUT are shown in Fig.~\ref{Fig:SliceNew}. The I/O signals are identical to the previous example in Fig.~\ref{Fig:ModifiedOA}. Normally 6-LUT can be configured with two different output ports \texttt{O6} and \texttt{O5}. For \texttt{LB1}, the carry input can be initialized by setting the \texttt{O6} of LUT2 to 0. In this case, the output of \texttt{MUXCY2} is always \texttt{O5} of LUT2, and the carry from \texttt{LB2} will not affect the result of \texttt{LB1}. For those logics to generate the MSD and the LSD, they can be combined into 1 Slice as observed from Fig.~\ref{Fig:ModifiedOA}. Therefore using this mapping method the resources within 1 Slice can be fully utilized, leading to a significant area reduction.

@@

\subsection{Performance Analysis}\label{subsec:OA_Performance}
In our experiments, we compare the proposed online adder (\texttt{OA\_new}) against the original design which is implemented using behavioural descriptions based on 4:2 compressors (\texttt{OA\_behv}).  In addition, we also include ripple carry adder (\texttt{RCA}) which uses conventional arithmetic, into the comparisons. All designs are targeting on Xilinx Virtex-6 FPGA with speed grade -1. Notice that in order to utilize the carry resources and to avoid logic optimizations by the synthesis tool, the Xilinx primitive component ``\texttt{Carry4}" is used to create \texttt{OA\_new}.

@@

We record the rated frequencies of all designs when varying the operand word-lengths. The results are shown in Fig.~\ref{Fig:AdderFreq}. The frequency values are obtained through Xilinx Timing Analyzer after placing and routing the designs in ISE 14.1. It can be seen that both designs with online arithmetic achieves relatively stable operating frequencies across a variety of operand word-lengths. This is as expected, because the critical path delay in the online adder is theoretically a fixed value and it is irrelevant to the operand precisions. In comparison, the rated frequency of the \texttt{RCA} drops drastically for larger word-lengths. It follows that both online adders outperform \texttt{RCA} for large word-lengths. Furthermore, our approach operates faster than \texttt{OA\_behv}; the speed-ups of \texttt{OA\_new} over \texttt{OA\_behv} are from $1.20\times$ to $1.26\times$.

We also compare the resource usages of all three adder designs with respect to a variety of operand word-lengths. The results are presented in Fig~\ref{Fig:AdderArea}. We can see that in comparison to the original \texttt{OA\_behv}, our implementation \texttt{OA\_new} achieves significant area savings: $25\%\sim 33\%$ in LUTs and $67\%\sim77\%$ in Slices. This is because the dedicated FPGA resources are fully utilized. However, our design still costs more area than an \texttt{RCA}, with an area overhead of $80\%\sim98\%$ in terms of LUTs, and up to $88\%$ in terms of Slices. This is because for a single adder, the area overhead brought by the redundant number representations cannot be removed.
%However research has shown that this overhead can be reduced for multiple-operand adders or adder trees [xxx].

@@

In comparison to Algorithm~\ref{Algorithm:OnlineMult}, the maximum number of stages required to generate results with full precision drops by a factor of 2 using the proposed new algorithm. In each stage, the word-length of all signals also reduces from $(N+2+\delta)$ to $(N+2)$, as $\delta$ is not necessary for digit parallel operations. In addition, the $frac()$ function can be implemented only based on wire connections without using other logic resources.

>>>>

\begin{abstract}
%\boldmath
Online arithmetic has been widely studied for ASIC implementation. Online components are designed to perform computations most significant digit first, resulting in the ability to chain arithmetic operators together for low latency. Recent research also shows that digit parallel online operators can fail more gracefully when operating beyond the deterministic clocking region, in comparison to operators with conventional arithmetic. Unfortunately, the utilization of online arithmetic operators in the past has required a large area overhead for FPGA implementation. In this paper, we propose novel approaches to implement the key primitives of online arithmetic, adders and multipliers, efficiently on modern Xilinx FPGAs with 6-input LUTs and carry resources. We demonstrate experimentally that in comparison to a direct implementation, the proposed architectures achieve over $67\%$ Slice savings and over $1.2\times$ speed-ups for adders, and more than $3\times$ area savings and $1.5\times$ speed-ups for multipliers.

@@

\section{Introduction}\label{Sec:Intro}
In the conventional form of computer arithmetic, the computation results are generated either from the least significant digit (LSD), e.g. addition and multiplication, or from the most significant digit (MSD), e.g. division and square root. This inconsistency in computing directions potentially results in large latency when propagating data between different operations. Online arithmetic was designed to solve this problem \cite{Ercegovac_OnlineOverview,Ercegovac_Book}. With online arithmetic, both inputs and outputs are processed in a MSD-first manner. This enables parallelism among multiple operations, and the overall computation latency can be significantly reduced. A brief overview of online arithmetic is given in Section~\ref{Sec:Background}.

Recent research has demonstrated another key advantage of online arithmetic: that it allows graceful degradation when operating beyond the deterministic clocking region, i.e. when overclocking \cite{SKDAC14_REVIEW}. This theme of research stems from the inspiration that for certain applications, releasing the absolute accuracy requirements could lead to significant performance improvements. However, traditional forms of computer arithmetic do not fail gracefully when timing violation happens, because timing errors tend to affect MSDs of the results. In comparison, online arithmetic is more ``overclocking friendly'', as timing errors only occur at LSDs of the results.

Despite its attractiveness, the use of online arithmetic on existing FPGAs is still limited due to a large area overhead. Given for standard arithmetic, it is difficult to efficiently map arithmetic circuits solely onto look-up-tables (LUTs), the fundamental building blocks of FPGAs. As a result, both main commercial FPGA vendors introduced dedicated carry logic into the logic block to reduce carry propagation delay~\cite{Virtex6}. Significant prior work explores methods that can effectively map circuits to the carry logic for performance improvements~\cite{FPL10_FPGA_CarryChain,Prefix_CarryChain}, or proposes alternative carry logic for FPGAs~\cite{FPGA_CarryChain_New1,FPGA_CarryChain_New2}. Hard DSP blocks are also included in modern architectures to improve the area and performance of multiplications and multiply accumulates~\cite{XilinxDSP}. However, most of these approaches are designed to accelerate computations with the conventional form of arithmetic.

For online arithmetic, two major research directions have been investigated. First, although there has been previous research into FPGA implementations of online arithmetic, this work only focused on online adders; a brief review of relevant works in this area is presented in Section~\ref{Sec:OA_FPGA}. Second, there are existing approaches to build general purpose online adders specifically for FPGAs with 4-input LUTs and with 2 LUTs within a Slice, such as the Altera Cyclone III \cite{AlteraCyclone} and Xilinx Spartan~3 series \cite{XilinxSpartan}. However, as will be discussed in Section~\ref{Sec:OA_FPGA}, these approaches cannot be directly applied on many modern FPGA with 6-input LUTs and 4 LUTs in a Slice, such as the Xilinx Virtex series FPGAs and all Xilinx 7 series FPGAs~\cite{Virtex7}. Moreover, the move to the 6-LUT on the basic architecture opens new opportunities for efficient implementation.
%However, the implementations of other key arithmetic primitives, such as online multipliers, lack exploration.

In this paper, both issues will be addressed. In Section~\ref{Sec:OA_FPGA}, we propose a novel approach to map the digit parallel online adders onto FPGAs with 6-LUTs. The mapping is based on the fast carry logic, and the available resources within a slice can be fully utilized. In Section~\ref{Sec:OM_FPGA}, we optimize the online multiplication algorithm to yield an efficient FPGA implementation of a digit parallel online multiplier. The resource usage of both operators can be determined analytically. We demonstrate experimentally on Xilinx Virtex-6 FPGAs that the proposed designs achieve significant area reduction and performance gain over their original implementations. For online adders, our method achieves over $67\%$ Slice savings and $1.2\times$ frequency speed-ups in comparison to a direct implementation. For online multipliers, the LUTs and Slice utilization of the proposed new architecture drops over $2.4\times$ and $3.2\times$, with a frequency speed-up more than $1.5\times$ in comparison to a direct implementation. In summary, the main contributions of this paper are an efficient mapping of online adders and multipliers to modern FPGAs. (to be completed)
% by fully utilizing the fast carry logic for .

@@

Online arithmetic has been widely used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.
%
\begin{figure}[tbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.pdf}
  \vspace{-4ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-1ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible only if a redundant number system is used. Normally there are two most commonly used redundant number representations: carry-save (CS) \cite{CSadder} and signed-digit (SD) \cite{RedundantNumber}. With SD representation, each digit is represented using a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. In comparison, the standard non-redundant representation only uses a digit set $\{0,\cdots,r-1\}$. Thus a standard number corresponds to several possible redundant representations. For example, the binary number $0.011$ can be represented in SD form as $0.1\overline{1}1$, $0.10\overline{1}$ or $0.011$ among many other possible representations.

Due to the redundancy, the MSDs of the result can be calculated using partial information from both inputs. Then the value of the number can be revised using the subsequent digits, because each number has multiple representations.

\subsection{Binary Online Addition}
Adders serve as a critical building block for arithmetic operations. To perform digit-parallel online addition, a redundant adder can be used directly. The structure of an online adder where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes 3 inputs and generates 2 outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labelled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only 2 FA delays for any operand word-length, with the cost of one extra FA for each digit of operands. This makes the online adder suitable for building up more complex arithmetic operators such as multipliers to accelerate the sum of partial products \cite{RedundantMult_1987}.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder2.pdf}
    \vspace{-6ex}
	\caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
    %\vspace{2ex}
	\label{Fig:Radix2SD_adder}
\end{figure}

@@

\subsection{Binary Online Multiplication}
Multiplication is another key arithmetic operator. Typically, online multiplication is performed in a recursive digit serial manner, as illustrated in Algorithm~\ref{Algorithm:OnlineMult} \cite{Ercegovac_OnlineMult} where both inputs and outputs are $N$-digit numbers as given in~(\ref{Eq:Online_Operands}). For a given iteration $j$, the product digit $z_j$ is generated MSD-first through a selection function $sel()$. For any radix $r$ and chosen digit set, there exits an appropriate selection method and a value of $\delta$ which ensure convergence \cite{Ercegovac_OnlineMult}. As radix 2 is used most commonly in computer arithmetic, we keep $r=2$ throughout this paper with the corresponding redundant digit set $\{\overline{1},0,1\}$. In this case $sel()$ is given by (\ref{Eq:SelFunc_OM}) \cite{Oregon_OnlineNetwork}. Notice that the selection is based on two integer digits and one fractional digit of $W_{[j]}$.

@@

It is worth noting that the precision of the product $Z$ can be determined by directly truncating from the LSDs, since there is no carry propagation from the LSD to the MSD. This is useful especially when the multiplication product is used by successive computations, where often only the first $N$ digits of the results are used. In comparison, the product is generated starting from the LSD with conventional arithmetic. Hence the successive operations cannot commence until the result with full-precision is generated. This feature of online arithmetic has inspired research about dynamically control of computation precision in numerous applications~\cite{Online_Truncation}.

@@

In Section~\ref{Sec:OM_FPGA}, we will first describe an optimized online multiplication algorithm and its FPGA implementation, which are designed specifically targeting digit parallel operations. Then we provide a modified structure which only generates the most significant half of the product in order to achieve extra area savings.

@@

\section{Digit Parallel Online Adder for FPGAs}\label{Sec:OA_FPGA}
\subsection{Related Works}
There have been previous works about FPGA implementation of digit parallel online adders. From the literature, the existing approaches can be classified into three types:

\begin{enumerate}
    \item efficient mapping of the digit-parallel online adder onto sophisticated FPGAs resources \cite{FPT09RA,ASAP09RA};
    \item multiple operand addition by designing compressor trees based on bit counters \cite{TC13RA,CSMult_CounterArray};
    \item modifying existing FPGA architecture for more efficient digit-parallel online addition \cite{DAC07_CounterTree} and for specific applications \cite{FPL09_onlineCCM}.
\end{enumerate}

Type 2 examines the use of counters to sum $k$ $N$-digit numbers where $k>2$. Instead, we focus on two-operand online addition that performs $Z=X+Y$, as shown in Fig.~\ref{Fig:Radix2SD_adder}. In comparison to type 3, this paper attempts to obtain the best performance from mainstream FPGAs instead of suggesting ways to modify the FPGA fabric.

Specifically in type 1, both works took advantage of the built-in carry resources in FPGAs. Conventionally the ASIC implementation of online adders is based on 4:2 compressors, as outlined in Fig.~\ref{Fig:PreviousWork} within the gray background. However, directly applying this approach in the FPGAs could be less efficient. This is because there is no carry propagation between the 2 FAs within a 4:2 compressor, and the net delay between them can be large. Instead, Kamp et al \cite{FPT09RA} and Hormigo et al \cite{ASAP09RA} described very similar mapping techniques for online adders with SD representations and CS representations, respectively. They share the common idea of using a maps the logic block (\texttt{LB}) within the dotted circle in Fig.~\ref{Fig:PreviousWork}. In this case, the fast-carry logic in the FPGA can be employed, and the delay between the two FAs can be greatly reduced.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder_42comp_Spartan2.pdf}
	\caption{Map the online adder onto Spartan FPGAs using the fast-carry resources. The gray background highlights the 4:2 compressor. Dotted circle indicates the logic block (LB) which can be mapped to the FPGA using the carry resources.}
	\label{Fig:PreviousWork}
\end{figure}

However, we notice the major limitations of both approaches that they only target FPGAs with 4-input LUTs (4-LUT) and two LUTs within a logic slice, such as the Xilinx Spartan series and the Altera Cyclone series. This is naturally reasonable because one \texttt{LB} can be mapped to a single Slice. Nevertheless, for FPGAs with 6-input LUTs (6-LUT) and four LUTs in a slice, such as the Xilinx Virtex series and all Xilinx 7 series FPGAs, directly applying these approaches will result in either resource waste or wrong results. For instance, if two \texttt{LB}s are mapped to a slice with four LUTs as seen in Fig.~\ref{Fig:PreWorkVirtex6}, the outputs of \texttt{LB1} should not be used because the its carry input cannot be explicitly initialized. Due to the same reason, the output of \texttt{XORCY2} should not be used.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=.45\textwidth]{./Figures/SDadder_FastCarry_PreviousWork2.pdf}
	\caption{An example illustrating that direct applying approaches in \cite{FPT09RA} on a Virtex-6 FPGA will results in faulty outputs, because the carry input of \texttt{LB1} cannot be explicitly initialized.}
	\label{Fig:PreWorkVirtex6}
    %\vspace{-2ex}
\end{figure}

\subsection{Proposed Mapping Method}
To tackle this problem, we first present an alternative structure of the online adder to enable an efficient FPGA mapping. The structure of a 4-digit online adder is given as an example in Fig.~\ref{Fig:ModifiedOA}. In this equivalent structure, the first FA in each 4:2 compressor is split into 2 parts, which only generate carry and sum respectively. In this case they can be mapped individually on 2 LUTs. The second FA, which generates the outputs, is unchanged and can be implemented using the fast carry logic.

@@

\begin{figure*}[tbp]
	\centering
	\includegraphics[width=.85\textwidth]{./Figures/SDadder_FPGA_New_4digit2.pdf}
	\caption{Alternative structure of online adder. Left: an example of 4-digit online adder. The shaded part refers to the 2 logic blocks (LBs) that can be mapped onto 1 Slice. Right: one LB. The dotted boxes outline the logic that can be mapped onto each LUT and the corresponding fast carry logic.}
	\vspace{-1ex}	
	\label{Fig:ModifiedOA}
\end{figure*}

The detailed slice mapping of the two \texttt{LB}s and the logic in each LUT are shown in Fig.~\ref{Fig:SliceNew}. The I/O signals are identical to the previous example in Fig.~\ref{Fig:ModifiedOA}. The 6-LUT can be configured with two different output ports \texttt{O6} and \texttt{O5}. For \texttt{LB1}, the carry input can be initialized by setting the \texttt{O6} of LUT2 to 0. In this case, the output of \texttt{MUXCY2} is always \texttt{O5} of LUT2, and the carry from \texttt{LB2} will not affect the result of \texttt{LB1}. The logic to generate the MSD and the LSD can be combined into one slice as observed from Fig.~\ref{Fig:ModifiedOA}. Therefore using this mapping method the resources within one slice can be fully utilized, leading to a significant area reduction.

@@

\subsection{Performance Analysis}\label{subsec:OA_Performance}
In our experiments, we compare the proposed online adder (\texttt{OA\_new}) against the original design which is implemented using functional descriptions in Verilog based on 4:2 compressors (\texttt{OA\_behv}).  In addition, we also include ripple carry adder (\texttt{RCA}) which uses conventional arithmetic, into the comparisons. All designs are targeting on Xilinx Virtex-6 FPGA with speed grade -1. Notice that in order to utilize the carry resources and to avoid logic optimizations by the synthesis tool, the Xilinx primitive component ``\texttt{Carry4}" is used to create \texttt{OA\_new}.

@@

We record the rated frequencies of all designs when varying the operand word-lengths. The results are shown in Fig.~\ref{Fig:AdderFreq}. The frequency values are obtained through Xilinx Timing Analyzer after placing and routing the designs in ISE 14.1. It can be seen that both designs with online arithmetic achieves relatively stable operating frequencies across a variety of operand word-lengths. This is as expected, because the critical path delay in the online adder is theoretically a fixed value independent of the operand precision. In comparison, the rated frequency of the \texttt{RCA} drops dramatically for larger word-lengths. It follows that both online adders outperform \texttt{RCA} for large word-lengths. Furthermore, our approach operates faster than \texttt{OA\_behv}; the speed-ups of \texttt{OA\_new} over \texttt{OA\_behv} are from $1.20\times$ to $1.26\times$.

We also compare the resource usage of all three adder designs with respect to a variety of operand word-lengths. The results are presented in Fig~\ref{Fig:AdderArea}. We can see that in comparison to the original \texttt{OA\_behv}, our implementation \texttt{OA\_new} achieves significant area savings: $25\%\sim 33\%$ in LUTs and $67\%\sim77\%$ in slices. This is because the dedicated FPGA resources are fully utilized. However, our design still costs more area than an \texttt{RCA}, with an area overhead of $80\%\sim98\%$ in terms of LUTs, and up to $88\%$ in terms of slices. This is because for a single adder, the area overhead brought by the redundant number representations cannot be removed.
%However research has shown that this overhead can be reduced for multiple-operand adders or adder trees [xxx].

@@

In comparison to Algorithm~\ref{Algorithm:OnlineMult}, the maximum number of stages required to generate results with full precision drops by a factor of 2 using the proposed new algorithm. In each stage, the word-length of all signals also reduces from $(N+2+\delta)$ to $(N+2)$, as $\delta$ is not necessary for digit parallel operations. In addition, the $frac()$ function can be implemented only based on wire connections without using logic resources.
